---
title: "Assignment 3"
author: "Group 4"
date: "2024-10-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
```{r imports, results='hide', message=FALSE}
options(contrasts = c("contr.sum", "contr.poly"))
require("ggplot2")
require("dplyr")
require("ppcor")
require("caret")
require("tidyr")
require("stringr")
require("lubridate")
require("tsibble")
require("ggfortify")
require("gridExtra")
require("reshape2")

library(ggcorrplot) # Easy cross-corrleation plots
library(imputeTS)   # Time series missing value imputation

library(jsonlite)   # handle JSON data returned by Frost
library(tidyr)      # unpack data from JSON format
library(tidyverse)  # data manipulation with mutate etc, string formatting
library(lubridate)  # process date and time information
library(tsibble)    # special tibbles for time series
library(fpp3)       # autoplot() and gg_season() for time series
library(readr)      # to read the Frost client ID from file

library(readxl)
```

# Part 1

## Task A

Faults, adjusting clock summer (+02:00) time winter time (+01:00) 

handle it by setting UTC as standard to transform to in the ymd_hms function.

```{r}
power_consum_raw <- read.table("data/consumption_per_group_aas_hour.csv", sep=";", dec=",", header= T)
colnames(power_consum_raw)

power_consum_df <- power_consum_raw  %>%
  dplyr::select(c("STARTTID", "FORBRUKSGRUPPE", "VOLUM_KWH")) %>%
  mutate(STARTTID = ymd_hms(STARTTID, tz = "UTC"))

summary(power_consum_df) 
head(power_consum_df)
```

```{r}
power_consum_df_long <- power_consum_df %>% 
  pivot_wider(names_from = FORBRUKSGRUPPE, values_from = VOLUM_KWH)

data_range <- seq(min(power_consum_df_long$STARTTID), max(power_consum_df_long$STARTTID), by = "1 hour")

last_date_gap <- tail(data_range[!data_range %in% power_consum_df_long$STARTTID], 1)

power_consum_df_long_cut <- power_consum_df_long %>%
  dplyr::filter(STARTTID >= last_date_gap)

head(power_consum_df_long_cut)
```

```{r}
head(power_consum_df_long_cut)
tail(power_consum_df_long_cut)
```


```{r}
# Mannualy remove the head and tail that don't sum to 1 whole day

power_consum_df_long_cut_cutDay <- power_consum_df_long_cut %>%
  dplyr::filter(STARTTID > ymd_hms("2021-04-30 23:00:00 UTC")) %>%
  dplyr::filter(STARTTID < ymd_hms("2024-09-30 00:00:00 UTC"))
head(power_consum_df_long_cut_cutDay)
tail(power_consum_df_long_cut_cutDay, 24)
```




```{r}
power_consum_sum_df <- power_consum_df_long_cut_cutDay %>%
  mutate(STARTTID = as.Date(STARTTID)) %>%
  group_by(STARTTID) %>%
  summarise_each(funs(sum))

head(power_consum_sum_df)
```
### Final clean

```{r}
power_consum_df_c <- power_consum_sum_df
```


## Task B

Data is daily. 

Data registered on different times, might skew results.

In the excell file the dates are stored in this format '01/01/2020 22:10:00'

Relevant years: 2017-2024:

* 

```{r}
file_path <- "data"
excel_files <- list.files(path = file_path, pattern = "*.xlsx", full.names = TRUE)
```
Data formats that might be problematic, some non unique dato. Leap years.
FIles DATE formated differently, might cause issues if the read_excel function does not account for it.



```{r}
data_range <- seq(min(metro_aas_df$DATO), max(metro_aas_df$DATO), by = "1 day")

metro_aas_df <- excel_files %>%
  lapply(read_excel) %>%
  bind_rows() %>%
  dplyr::select(c("DATO","LT", "GLOB")) %>%
  mutate(DATO = as.Date(DATO)) %>%
  group_by(DATO) %>%
  distinct(DATO, .keep_all = TRUE) %>%
  ungroup() %>%
  complete(DATO = seq(min(DATO), max(DATO), by = 'day')) # Add missing days

summary(metro_aas_df)
head(metro_aas_df)
max(metro_aas_df$DATO)
```


### Missing data analysis

```{r}
metro_aas_df_c <- metro_aas_df
```




```{r}
ggplot_na_distribution(metro_aas_df_c$LT)
```

```{r}
ggplot_na_distribution(metro_aas_df_c$GLOB)
```

As seen in the previous graphs, there are very few missing values and they seem evenly spread out, therfore imputations seems to be a good choice to fill in the missing days.

```{r}
metro_aas_df_imputed <- imputeTS::na_ma(metro_aas_df_c , k = 4, weighting = "simple")
head(metro_aas_df_imputed)
```

```{r}
metro_aas_df_imputed %>%
  ggplot(aes(DATO, LT)) +
  geom_line()

metro_aas_df_imputed %>%
  ggplot(aes(DATO, GLOB)) +
  geom_line()
```
### Final cleaned data

```{r}
metro_aas_df_c <- metro_aas_df_imputed
```

## Task C

### Find the range of dates for each of the two data sets.

```{r}
power_consum_dateRange <- c(min(power_consum_df_c$STARTTID), max(power_consum_df_c$STARTTID))
metro_ass_dateRange <- c(min(metro_aas_df_c$DATO), max(metro_aas_df_c$DATO))
power_consum_dateRange
metro_ass_dateRange
```
### For the longest contiguous range of dates present in both data sets, merge the two data sets based on date.

```{r}
merged_df <- dplyr::inner_join(power_consum_df_c, metro_aas_df_c, by = join_by(STARTTID == DATO)) %>%
  rename(DATO = STARTTID)
  
head(merged_df)
```

### Remove data for leap days.

```{r}
merged_df_noLeap <- merged_df %>% 
  dplyr::filter(!(month(DATO) == 2 & day(DATO) == 29))

head(merged_df)
```




